---
title: "DENV major-lineage sampling through time"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(stringr)
library(ggplot2)
library(scales)
library(tibble)
library(colorspace)
library(patchwork)
library(purrr)
library(forcats)
library(grid)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

serotype_levels <- c("DENV1", "DENV2", "DENV3", "DENV4")
serotype_labels <- c(
  "DENV1" = "DENV-1",
  "DENV2" = "DENV-2",
  "DENV3" = "DENV-3",
  "DENV4" = "DENV-4"
)

dataset_levels <- c("Full dataset", "Subsampled dataset")

# Set to 1 to show every major lineage. Keep at 5 to collapse very rare lineages.
lineage_min_n <- 5
lineage_other_label <- paste0("Other lineages (<", lineage_min_n, ")")
other_color <- "grey72"
unknown_color <- "grey45"

figure_output_dir <- "~/Desktop/DENV_EVO_2025/Subsampling_figures_major_lineage"
dir.create(figure_output_dir, recursive = TRUE, showWarnings = FALSE)

base_sampling_theme <- theme_classic(base_size = 9) +
  theme(
    axis.title = element_text(face = "bold", size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 12),
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.text = element_text(size = 10),
    legend.key.size = unit(0.13, "cm"),
    legend.spacing.x = unit(0.04, "cm"),
    legend.spacing.y = unit(0.00, "cm"),
    legend.margin = margin(-4, 0, -4, 0),
    legend.box.margin = margin(-6, 0, -6, 0),
    plot.margin = margin(4, 5, 4, 5)
  )
```

# File paths

Update `metadata_dir` if needed. The import function accepts `.csv`, `.tsv`, and `.txt` files.

```{r file-paths}
metadata_dir <- "~/Desktop/DENV_EVO_2025/Lineage_metadata"

metadata_files <- tibble::tribble(
  ~serotype, ~dataset, ~file_stem,
  "DENV1", "Full dataset",       "DENV1_together",
  "DENV2", "Full dataset",       "DENV2_together",
  "DENV3", "Full dataset",       "DENV3_together",
  "DENV4", "Full dataset",       "DENV4_together",
  "DENV1", "Subsampled dataset", "DENV1_sampled",
  "DENV2", "Subsampled dataset", "DENV2_sampled",
  "DENV3", "Subsampled dataset", "DENV3_sampled",
  "DENV4", "Subsampled dataset", "DENV4_sampled"
) %>%
  mutate(file_path = file.path(metadata_dir, file_stem))
```

# Helper functions

```{r helpers}
resolve_metadata_path <- function(file_stem_path) {
  candidates <- c(
    paste0(file_stem_path, ".csv"),
    paste0(file_stem_path, ".tsv"),
    paste0(file_stem_path, ".txt")
  )
  hit <- candidates[file.exists(path.expand(candidates))]
  if (length(hit) == 0) {
    stop("Could not find metadata file for: ", file_stem_path,
         "\nChecked: ", paste(candidates, collapse = ", "))
  }
  hit[[1]]
}

read_delim_auto <- function(file_path) {
  resolved <- resolve_metadata_path(file_path)
  ext <- tools::file_ext(resolved)
  if (tolower(ext) == "tsv" || tolower(ext) == "txt") {
    read_tsv(resolved, show_col_types = FALSE)
  } else {
    read_csv(resolved, show_col_types = FALSE)
  }
}

find_first_existing_col <- function(df, candidates) {
  hit <- candidates[candidates %in% names(df)]
  if (length(hit) == 0) NA_character_ else hit[[1]]
}

extract_decimal_year <- function(x) {
  if (inherits(x, "Date")) {
    yr <- as.numeric(format(x, "%Y"))
    doy <- as.numeric(format(x, "%j"))
    return(yr + (doy - 1) / 365.25)
  }

  x_chr <- str_trim(as.character(x))
  x_chr[x_chr %in% c("", "NA", "N/A", "na", "unknown", "Unknown", "?")] <- NA_character_
  x_num <- suppressWarnings(as.numeric(x_chr))
  out <- rep(NA_real_, length(x_chr))

  decimal_or_plain_year <- !is.na(x_num) & x_num >= 1900 & x_num <= 2100
  out[decimal_or_plain_year] <- x_num[decimal_or_plain_year]

  excel_serial <- !is.na(x_num) & x_num > 20000 & x_num < 80000
  if (any(excel_serial)) {
    d <- as.Date(x_num[excel_serial], origin = "1899-12-30")
    yr <- as.numeric(format(d, "%Y"))
    doy <- as.numeric(format(d, "%j"))
    out[excel_serial] <- yr + (doy - 1) / 365.25
  }

  still_missing <- is.na(out) & !is.na(x_chr)
  date_formats <- c(
    "%Y-%m-%d", "%Y/%m/%d", "%Y.%m.%d",
    "%m/%d/%Y", "%m-%d-%Y", "%m.%d.%Y",
    "%d/%m/%Y", "%d-%m-%Y", "%d.%m.%Y",
    "%Y-%m", "%Y/%m", "%Y.%m"
  )

  for (fmt in date_formats) {
    idx <- which(still_missing)
    if (length(idx) == 0) break
    d <- suppressWarnings(as.Date(x_chr[idx], format = fmt))
    good <- !is.na(d)
    if (any(good)) {
      yr <- as.numeric(format(d[good], "%Y"))
      doy <- as.numeric(format(d[good], "%j"))
      out[idx[good]] <- yr + (doy - 1) / 365.25
    }
    still_missing <- is.na(out) & !is.na(x_chr)
  }

  still_missing <- is.na(out) & !is.na(x_chr)
  fallback_year <- suppressWarnings(as.numeric(str_extract(x_chr[still_missing], "(19|20)\\d{2}")))
  out[still_missing] <- fallback_year
  out
}

make_shared_breaks <- function(years, n = 10) {
  years <- sort(unique(years[is.finite(years)]))
  if (length(years) <= n) return(years)
  pretty(range(years), n = n)
}

# Major lineage strings look like 1I_K, 1III_A, 2V_B, 4II_A.
# This extracts the part before the underscore so that lineages from the
# same genotype/major group get related shades.
extract_lineage_group <- function(x) {
  x <- as.character(x)
  out <- str_extract(x, "^[^_]+")
  out[is.na(out) | out == ""] <- x[is.na(out) | out == ""]
  out
}

# Palette rule:
# - each group before underscore gets one base hue
# - individual lineages within the same group get related DARK shades
# - no near-white shades, because pale colors make stacked bars look patchy
# - Other and Unknown remain greys
make_grouped_lineage_palette <- function(lineage_levels, other_name = lineage_other_label) {
  lineage_levels <- as.character(lineage_levels)
  lineage_levels <- lineage_levels[!is.na(lineage_levels)]

  special_levels <- lineage_levels[lineage_levels %in% c(other_name, "Unknown lineage", "unassigned", "Unassigned")]
  main_levels <- setdiff(lineage_levels, special_levels)

  if (length(main_levels) == 0) {
    cols <- character(0)
  } else {
    lineage_groups <- extract_lineage_group(main_levels)
    group_levels <- unique(lineage_groups)

    # Dark, saturated base colors. These are intentionally not pastel.
    base_pool <- c(
      "#B2182B", # red
      "#2166AC", # blue
      "#1B7837", # green
      "#762A83", # purple
      "#B35806", # brown/orange
      "#008837", # teal green
      "#C51B7D", # magenta
      "#5E3C99", # indigo
      "#01665E", # dark teal
      "#8C510A"  # dark brown
    )
    base_cols <- rep(base_pool, length.out = length(group_levels))
    names(base_cols) <- group_levels

    cols <- unlist(lapply(group_levels, function(g) {
      levs <- main_levels[lineage_groups == g]
      levs <- levs[order(levs)]
      n <- length(levs)
      base_col <- base_cols[[g]]

      if (n == 1) {
        shade_cols <- base_col
      } else {
        # Dark to medium variants only. This avoids white/pastel bars.
        shade_steps <- seq(0.38, -0.18, length.out = n)
        shade_cols <- vapply(shade_steps, function(s) {
          if (s >= 0) {
            colorspace::darken(base_col, amount = s)
          } else {
            colorspace::lighten(base_col, amount = abs(s))
          }
        }, character(1))
      }

      setNames(shade_cols, levs)
    }))
  }

  if (other_name %in% lineage_levels) {
    cols <- c(cols, setNames("grey55", other_name))
  }
  if ("Unknown lineage" %in% lineage_levels) {
    cols <- c(cols, setNames("grey30", "Unknown lineage"))
  }
  if ("unassigned" %in% lineage_levels) {
    cols <- c(cols, setNames("grey30", "unassigned"))
  }
  if ("Unassigned" %in% lineage_levels) {
    cols <- c(cols, setNames("grey30", "Unassigned"))
  }

  cols[lineage_levels]
}

read_metadata_file <- function(file_path, serotype_label, dataset_label) {
  df <- read_delim_auto(file_path)

  lineage_col <- find_first_existing_col(df, c("major_lineage", "Major_Lineage", "Major lineage", "lineage", "Lineage"))
  genotype_col <- find_first_existing_col(df, c("genotype", "Genotype", "GENOTYPE"))
  date_col <- find_first_existing_col(df, c(
    "date", "Date", "collection_date", "Collection_Date", "Collection date",
    "sampling_date", "Sampling_Date", "sample_date", "Sample_Date", "year", "Year"
  ))
  country_col <- find_first_existing_col(df, c(
    "country", "Country", "location", "Location", "country_exposure", "Country_exposure"
  ))
  id_col <- find_first_existing_col(df, c(
    "strain", "Strain", "name", "Name", "sequence_name", "Sequence_name", "accession", "Accession", "genbank_accession"
  ))

  if (is.na(lineage_col)) df <- df %>% mutate(major_lineage = NA_character_) else df <- df %>% mutate(major_lineage = as.character(.data[[lineage_col]]))
  if (is.na(genotype_col)) df <- df %>% mutate(genotype = NA_character_) else df <- df %>% mutate(genotype = as.character(.data[[genotype_col]]))
  if (is.na(country_col)) df <- df %>% mutate(country = NA_character_) else df <- df %>% mutate(country = as.character(.data[[country_col]]))
  if (is.na(id_col)) df <- df %>% mutate(sequence_id = NA_character_) else df <- df %>% mutate(sequence_id = as.character(.data[[id_col]]))
  if (is.na(date_col)) stop(paste0("No date/year column found in ", file_path))

  df %>%
    mutate(
      serotype = serotype_label,
      dataset = dataset_label,
      major_lineage = str_trim(major_lineage),
      major_lineage = if_else(is.na(major_lineage) | major_lineage == "", "Unknown lineage", major_lineage),
      sample_year = extract_decimal_year(.data[[date_col]]),
      year = floor(sample_year)
    ) %>%
    select(dataset, serotype, sequence_id, country, genotype, major_lineage, sample_year, year, everything())
}
```

# Import metadata

```{r import-metadata}
metadata_df <- purrr::pmap_dfr(
  metadata_files,
  function(serotype, dataset, file_stem, file_path) {
    read_metadata_file(file_path, serotype, dataset)
  }
) %>%
  mutate(
    dataset = factor(dataset, levels = dataset_levels),
    serotype = factor(serotype, levels = serotype_levels)
  ) %>%
  filter(!is.na(year))

metadata_df %>%
  count(serotype, dataset, name = "n_sequences") %>%
  arrange(serotype, dataset) %>%
  knitr::kable()
```

# Diagnostic checks for impossible or suspicious subsampling detections

These checks are meant to catch cases where the subsampled metadata contains bins that are not represented in the full metadata, or where a subsampled bin has more genomes than the corresponding full bin. A few early-year bars can look more visible in the subsampled panel because the y-axis is smaller, but they should still be present in the full dataset.

```{r subsampling-diagnostics}
full_bins <- metadata_df %>%
  filter(dataset == "Full dataset") %>%
  count(serotype, year, major_lineage, name = "full_n")

sampled_bins <- metadata_df %>%
  filter(dataset == "Subsampled dataset") %>%
  count(serotype, year, major_lineage, name = "sampled_n")

subsampled_bin_check <- sampled_bins %>%
  full_join(full_bins, by = c("serotype", "year", "major_lineage")) %>%
  mutate(
    full_n = dplyr::coalesce(full_n, 0L),
    sampled_n = dplyr::coalesce(sampled_n, 0L),
    issue = case_when(
      sampled_n > 0 & full_n == 0 ~ "Subsampled bin absent from full dataset",
      sampled_n > full_n ~ "Subsampled bin count exceeds full dataset",
      TRUE ~ "OK"
    )
  ) %>%
  filter(issue != "OK") %>%
  arrange(serotype, year, major_lineage)

if (nrow(subsampled_bin_check) == 0) {
  cat("No impossible serotype-year-lineage bins detected: every subsampled bin is present in the full dataset and sampled counts do not exceed full counts.\n")
} else {
  knitr::kable(subsampled_bin_check)
}

# Extra check for early-year detections that may visually appear only in the subsampled panel.
early_year_cutoff <- 1990

early_year_check <- sampled_bins %>%
  filter(year < early_year_cutoff) %>%
  left_join(full_bins, by = c("serotype", "year", "major_lineage")) %>%
  mutate(full_n = dplyr::coalesce(full_n, 0L)) %>%
  arrange(serotype, year, major_lineage)

cat("\nEarly-year subsampled bins before", early_year_cutoff, ". These should also have full_n > 0.\n\n")
knitr::kable(early_year_check)
```

# Prepare major-lineage groups

Lineages with fewer than `r lineage_min_n` genomes in the full dataset for that serotype are collapsed into `Other lineages`. Set `lineage_min_n <- 1` in the setup chunk to show every lineage.

```{r prepare-lineage-data}
lineage_keep <- metadata_df %>%
  filter(dataset == "Full dataset", !is.na(major_lineage), major_lineage != "", major_lineage != "Unknown lineage") %>%
  count(serotype, major_lineage, name = "total_n") %>%
  filter(total_n >= lineage_min_n)

lineage_plot_df <- metadata_df %>%
  filter(!is.na(major_lineage), major_lineage != "", major_lineage != "Unknown lineage") %>%
  left_join(lineage_keep %>% mutate(keep_lineage = TRUE), by = c("serotype", "major_lineage")) %>%
  mutate(
    lineage_plot = if_else(!is.na(keep_lineage) & keep_lineage, major_lineage, lineage_other_label),
    lineage_group = extract_lineage_group(lineage_plot),
    serotype_label = recode(as.character(serotype), !!!serotype_labels),
    panel = paste0(serotype_label, " | ", dataset),
    panel = factor(
      panel,
      levels = as.vector(t(outer(unname(serotype_labels[serotype_levels]), dataset_levels, paste, sep = " | ")))
    )
  ) %>%
  count(serotype, serotype_label, dataset, panel, year, lineage_plot, lineage_group, name = "n")

lineage_levels <- lineage_plot_df %>%
  group_by(lineage_plot) %>%
  summarise(total_n = sum(n), .groups = "drop") %>%
  arrange(desc(total_n)) %>%
  pull(lineage_plot)
lineage_levels <- c(setdiff(lineage_levels, lineage_other_label), intersect(lineage_other_label, lineage_levels))

year_breaks <- make_shared_breaks(lineage_plot_df$year, n = 10)

lineage_plot_df <- lineage_plot_df %>%
  mutate(lineage_plot = factor(lineage_plot, levels = lineage_levels))

lineage_plot_df %>%
  count(serotype_label, dataset, name = "n_year_lineage_bins") %>%
  knitr::kable()
```

# Figure: temporal sampling counts colored by major lineage

Each serotype is plotted as a paired full-versus-subsampled panel. Within each serotype, lineages from the same genotype group use related shades.

```{r major-lineage-count-figure, fig.width=8.27, fig.height=11.69}
make_serotype_lineage_plot <- function(st) {
  st_df <- lineage_plot_df %>%
    filter(serotype == st) %>%
    mutate(
      dataset = factor(as.character(dataset), levels = dataset_levels),
      lineage_plot = droplevels(lineage_plot)
    )

  st_lineage_levels <- st_df %>%
    group_by(lineage_plot) %>%
    summarise(total_n = sum(n), .groups = "drop") %>%
    mutate(
      lineage_group = extract_lineage_group(as.character(lineage_plot)),
      is_other = as.character(lineage_plot) == lineage_other_label
    ) %>%
    arrange(is_other, lineage_group, desc(total_n), as.character(lineage_plot)) %>%
    pull(lineage_plot) %>%
    as.character()

  st_colors <- make_grouped_lineage_palette(st_lineage_levels, other_name = lineage_other_label)
  st_breaks <- make_shared_breaks(st_df$year, n = 8)
  st_label <- serotype_labels[[as.character(st)]]

  st_df <- st_df %>%
    mutate(lineage_plot = factor(as.character(lineage_plot), levels = st_lineage_levels))

  ggplot(st_df, aes(x = year, y = n, fill = lineage_plot)) +
    geom_col(width = 0.95, color = NA) +
    facet_wrap(~ dataset, ncol = 2, scales = "free_y") +
    scale_fill_manual(values = st_colors, drop = FALSE) +
    scale_x_continuous(
      breaks = st_breaks,
      expand = expansion(mult = c(0.005, 0.01))
    ) +
    scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.10))) +
    labs(
      title = st_label,
      x = "Collection year",
      y = "Number of genomes"
    ) +
    base_sampling_theme +
    theme(
      plot.title = element_text(face = "bold", size = 12, hjust = 0),
      legend.position = c(0.015, 0.965),
      legend.justification = c(0, 1),
      legend.direction = "vertical",
      legend.background = element_rect(fill = scales::alpha("white", 0.78), color = NA),
      legend.text = element_text(size = 10),
      legend.key.size = unit(0.11, "cm"),
      legend.spacing.x = unit(0.03, "cm"),
      legend.spacing.y = unit(-0.02, "cm"),
      legend.margin = margin(1, 1, 1, 1),
      plot.margin = margin(2, 5, 2, 5)
    ) +
    guides(
      fill = guide_legend(
        ncol = 3,
        byrow = FALSE,
        override.aes = list(linewidth = 0, size = 1.8)
      )
    )
}

serotype_lineage_plots <- lapply(serotype_levels, make_serotype_lineage_plot)

p_major_lineage_counts <- wrap_plots(serotype_lineage_plots, ncol = 1)

p_major_lineage_counts
```

# Save figure

```{r save-major-lineage-count-figure}
png_path <- file.path(figure_output_dir, "DENV_all_serotypes_full_vs_subsampled_major_lineage_counts_by_year_grouped_shades.png")
pdf_path <- file.path(figure_output_dir, "DENV_all_serotypes_full_vs_subsampled_major_lineage_counts_by_year_grouped_shades.pdf")

ggsave(png_path, p_major_lineage_counts, width = 8.27, height = 11.69, dpi = 600)
ggsave(pdf_path, p_major_lineage_counts, width = 8.27, height = 11.69)

tibble(output = c("PNG", "PDF"), path = c(png_path, pdf_path)) %>%
  knitr::kable()
```

# Collapse check

```{r collapse-check}
lineage_collapse_check <- metadata_df %>%
  filter(dataset == "Full dataset", !is.na(major_lineage), major_lineage != "", major_lineage != "Unknown lineage") %>%
  count(serotype, major_lineage, name = "total_n") %>%
  mutate(plot_group = if_else(total_n >= lineage_min_n, major_lineage, lineage_other_label)) %>%
  count(serotype, plot_group, name = "n_lineages") %>%
  filter(plot_group == lineage_other_label)

cat("Lineage groups collapsed using strict n <", lineage_min_n, "within the full dataset for each serotype.\n\n")
knitr::kable(lineage_collapse_check)
```
