---
title: "DENV selection site barcode plot"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Load packages

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(scales)
```

## Read and combine datasets

```{r}
files <- c(
  "DENV1_combined_sites.csv",
  "DENV2_combined_sites.csv",
  "DENV3_combined_sites.csv",
  "DENV4_combined_sites.csv"
)

serotypes <- c("DENV1", "DENV2", "DENV3", "DENV4")

df_list <- lapply(seq_along(files), function(i) {
  read_csv(files[i], show_col_types = FALSE) %>%
    mutate(serotype = serotypes[i])
})

all_sites <- bind_rows(df_list)
```

## Inspect gene names

```{r}
unique(all_sites$gene)
```

## Clean gene labels

```{r}
all_sites <- all_sites %>%
  mutate(
    gene_clean = case_when(
      grepl("capsid", gene, ignore.case = TRUE) ~ "C",
      grepl("prM|membrane", gene, ignore.case = TRUE) ~ "prM",
      grepl("envelope", gene, ignore.case = TRUE) ~ "E",
      grepl("NS1", gene, ignore.case = TRUE) ~ "NS1",
      grepl("NS2A", gene, ignore.case = TRUE) ~ "NS2A",
      grepl("NS2B", gene, ignore.case = TRUE) ~ "NS2B",
      grepl("NS3", gene, ignore.case = TRUE) ~ "NS3",
      grepl("NS4A", gene, ignore.case = TRUE) ~ "NS4A",
      grepl("NS4B", gene, ignore.case = TRUE) ~ "NS4B",
      grepl("NS5|polymerase", gene, ignore.case = TRUE) ~ "NS5",
      TRUE ~ gene
    )
  )

gene_order <- c("C", "prM", "E", "NS1", "NS2A", "NS2B", "NS3", "NS4A", "NS4B", "NS5")

all_sites <- all_sites %>%
  mutate(
    gene_clean = factor(gene_clean, levels = gene_order),
    serotype = factor(serotype, levels = c("DENV1", "DENV2", "DENV3", "DENV4"))
  )
```

## Build plotting dataframe

This version keeps significant FEL calls in one `FEL` column, using color to distinguish FEL- and FEL+. FEL neutral calls are removed because they are not significant selection sites. FEL+ is also duplicated into a separate adjacent `FEL+` column for emphasis, but both FEL+ displays use the same legend category and color.

```{r}
fel_all <- all_sites %>%
  filter(fel_selection %in% c("negative", "positive")) %>%
  transmute(
    serotype,
    gene_clean,
    site,
    method = "FEL",
    signal = case_when(
      fel_selection == "negative" ~ "FEL-",
      fel_selection == "positive" ~ "FEL+"
    )
  )

fel_pos_focus <- all_sites %>%
  filter(fel_selection == "positive") %>%
  transmute(
    serotype,
    gene_clean,
    site,
    method = "FEL+",
    signal = "FEL+"
  )

meme_sig <- all_sites %>%
  filter(!is.na(meme_qval), meme_qval < 0.2) %>%
  transmute(
    serotype,
    gene_clean,
    site,
    method = "MEME",
    signal = "MEME"
  )

prime_sig <- all_sites %>%
  filter(!is.na(prime_marker), prime_marker != "-") %>%
  transmute(
    serotype,
    gene_clean,
    site,
    method = "PRIME",
    signal = "PRIME"
  )

plot_df <- bind_rows(fel_all, fel_pos_focus, meme_sig, prime_sig) %>%
  mutate(
    method = factor(method, levels = c("FEL", "FEL+", "MEME", "PRIME")),
    method_num = as.numeric(method),
    gene_clean = factor(gene_clean, levels = gene_order),
    serotype = factor(serotype, levels = c("DENV1", "DENV2", "DENV3", "DENV4"))
  )
```

## Check that data are present

```{r}
table(plot_df$method, plot_df$signal, useNA = "ifany")
head(plot_df)
```

## Axis breaks by protein

Capsid is short, so the y-axis is manually set to avoid crowding with 25 and 75.

```{r}
protein_breaks <- list(
  C    = c(1, 50, 100),
  prM  = c(1, 50, 100, 166),
  E    = c(1, 100, 200, 300, 400, 495),
  NS1  = c(1, 100, 200, 300, 352),
  NS2A = c(1, 50, 100, 150, 218),
  NS2B = c(1, 50, 130),
  NS3  = c(1, 200, 400, 618),
  NS4A = c(1, 50, 100, 150),
  NS4B = c(1, 100, 200, 248),
  NS5  = c(1, 300, 600, 900)
)

protein_scales <- lapply(names(protein_breaks), function(prot) {
  scale_y_reverse(
    breaks = protein_breaks[[prot]],
    name = "Site position"
  )
})
names(protein_scales) <- names(protein_breaks)
```

## Plot selection sites

```{r fig.width=13, fig.height=16}
# Install ggh4x if needed:
# install.packages("ggh4x")

if (!requireNamespace("ggh4x", quietly = TRUE)) {
  stop("Please install ggh4x first: install.packages('ggh4x')")
}

p <- ggplot(
  plot_df,
  aes(
    x = method_num,
    y = site,
    color = signal
  )
) +
  geom_segment(
    data = plot_df %>% filter(signal == "FEL+"),
    aes(
      x = method_num - 0.33,
      xend = method_num + 0.33,
      y = site,
      yend = site
    ),
    color = "black",
    linewidth = 0.85,
    inherit.aes = FALSE
  ) +
  geom_segment(
    aes(
      x = method_num - 0.32,
      xend = method_num + 0.32,
      y = site,
      yend = site
    ),
    linewidth = 0.65,
    lineend = "butt"
  ) +
  facet_grid(
    rows = vars(gene_clean),
    cols = vars(serotype),
    scales = "free_y",
    space = "free_y",
    switch = "y"
  ) +
  ggh4x::facetted_pos_scales(
    y = protein_scales
  ) +
  scale_x_continuous(
    breaks = seq_along(levels(plot_df$method)),
    labels = levels(plot_df$method),
    expand = expansion(mult = c(0.08, 0.08))
  ) +
  scale_color_manual(
    values = c(
      "FEL-" = "#0072B2",
      "FEL+" = "#D55E00",
      "MEME" = "#E69F00",
      "PRIME" = "#009E73"
    ),
    breaks = c("FEL-", "FEL+", "MEME", "PRIME"),
    labels = c("FEL negative", "FEL positive", "MEME", "PRIME")
  ) +
  labs(
    x = NULL,
    y = "Site position"
  ) +
  theme_bw(base_size = 18) +
  theme(
    panel.grid = element_blank(),
    strip.background = element_blank(),
    strip.placement = "outside",
    strip.text.x = element_text(face = "bold", size = 18),
    strip.text.y.left = element_text(face = "bold", size = 18, angle = 0),
    axis.text.x = element_text(size = 16, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 15),
    axis.title.y = element_text(size = 18),
    legend.title = element_blank(),
    legend.text = element_text(size = 15),
    legend.position = "right",
    panel.spacing.x = unit(0.7, "lines"),
    panel.spacing.y = unit(0.9, "lines")
  )

p
```

## Save plot

```{r}
ggsave(
  "DENV_selection_barcode_plot_flipped_FEL_single_legend.png",
  plot = p,
  width = 13,
  height = 16,
  dpi = 600
)

ggsave(
  "DENV_selection_barcode_plot_flipped_FEL_single_legend.pdf",
  plot = p,
  width = 13,
  height = 16,
  device = cairo_pdf
)
```

## Optional: individual protein plots

```{r eval=FALSE}
for (prot in gene_order) {
  prot_df <- plot_df %>%
    filter(gene_clean == prot)

  p_prot <- ggplot(
    prot_df,
    aes(x = method_num, y = site, color = signal)
  ) +
    geom_segment(
      data = prot_df %>% filter(signal == "FEL+"),
      aes(
        x = method_num - 0.33,
        xend = method_num + 0.33,
        y = site,
        yend = site
      ),
      color = "black",
      linewidth = 0.85,
      inherit.aes = FALSE
    ) +
    geom_segment(
      aes(
        x = method_num - 0.32,
        xend = method_num + 0.32,
        y = site,
        yend = site
      ),
      linewidth = 0.65,
      lineend = "butt"
    ) +
    facet_grid(
      rows = vars(gene_clean),
      cols = vars(serotype),
      scales = "free_y",
      space = "free_y",
      switch = "y"
    ) +
    scale_y_reverse(
      breaks = if (prot == "C") c(1, 50, 100) else waiver(),
      name = "Site position"
    ) +
    scale_x_continuous(
      breaks = seq_along(levels(plot_df$method)),
      labels = levels(plot_df$method),
      expand = expansion(mult = c(0.08, 0.08))
    ) +
    scale_color_manual(
      values = c(
        "FEL-" = "#0072B2",
        "FEL+" = "#D55E00",
        "MEME" = "#E69F00",
        "PRIME" = "#009E73"
      )
    ) +
    labs(x = NULL, y = "Site position") +
    theme_bw(base_size = 18) +
    theme(
      panel.grid = element_blank(),
      strip.background = element_blank(),
      strip.placement = "outside",
      strip.text.x = element_text(face = "bold", size = 18),
      strip.text.y.left = element_text(face = "bold", size = 18, angle = 0),
      axis.text.x = element_text(size = 16, angle = 45, hjust = 1),
      axis.text.y = element_text(size = 15),
      legend.title = element_blank(),
      legend.text = element_text(size = 15),
      legend.position = "right"
    )

  print(p_prot)

  ggsave(
    filename = paste0("Blownup_", prot, "_selection_plot_flipped_FEL_focus.png"),
    plot = p_prot,
    width = 10,
    height = 8,
    dpi = 600
  )
}
```
