---
title: "DENV Evolutionary Rate Analysis with Summary Plot"
author: "Afeez"
date: "`04-16-26`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
---

```{r setup, message=FALSE, warning=FALSE}
# install.packages(c("readxl", "dplyr", "tidyr", "stringr", "ggplot2", "forcats", "scales", "patchwork"))

library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(forcats)
library(scales)
library(patchwork)
# Use + theme_manuscript on all plots before combining

# Increase spacing between combined panels so x-axis labels
# from adjacent plots do not visually merge together.
# Example:
# combined_plot <- (plot1 / plot2 / plot3) +
#   plot_layout(heights = c(1, 1, 1)) +
#   plot_annotation(tag_levels = "a")

library(readr)
library(purrr)
library(tibble)

# Shared DENV serotype colors used across all evolutionary-rate plots
serotype_colors <- c(
  "DENV1" = "#227DCB",
  "DENV2" = "#b020e0",
  "DENV3" = "#ffce9d",
  "DENV4" = "#954226"
)

serotype_labels <- c(
  "DENV1" = "DENV-1",
  "DENV2" = "DENV-2",
  "DENV3" = "DENV-3",
  "DENV4" = "DENV-4"
)

# Log-scale axis used across evolutionary-rate plots.
# Starts at 1e-4 and includes intermediate ticks where these rate estimates cluster.
rate_log_breaks <- c(1e-4, 2e-4, 3e-4, 5e-4, 7e-4, 1e-3, 2e-3, 3e-3, 5e-3, 1e-2)
rate_log_labels <- scales::label_scientific(digits = 1)



# =========================================================
# Shared manuscript theme for all combined figures
# =========================================================

theme_manuscript <- theme_bw(base_size = 12) +
  theme(
    plot.title = element_blank(),

    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 11),

    strip.text = element_text(size = 11, face = "bold"),

    plot.tag = element_text(
      size = 15,
      face = "bold"
    ),

    plot.tag.position = c(0.01, 0.98),

    panel.grid.minor = element_blank(),

    plot.margin = margin(15, 10, 15, 10)
  )


knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This document parses BEAST/Tracer-style summary sheets from the workbook and performs a protein-level comparison of evolutionary rates across all four dengue serotypes. The analysis uses the following sheets:

- `DENV1_Proteins`
- `DENV2_Proteins`
- `DENV3_Proteins`
- `DENV4_Proteins`

The main figure compares the **mean evolutionary rate** and **95% HPD interval** for each protein:

- C
- P
- E
- NS1
- NS2A
- NS2B
- NS3
- NS4A
- NS4B
- NS5

# File path

Update this path if your Excel file is stored somewhere else.

```{r file-path}
file_path <- "~/Desktop/DENV_EVO_2025/2026.03.25_all_evo_rates.xlsx"
```

# Helper functions

```{r helpers}
# Parse one BEAST/Tracer-style sheet
parse_rate_sheet <- function(file_path, sheet_name, group_col_name) {
  raw <- read_excel(file_path, sheet = sheet_name, col_names = TRUE)
  names(raw)[1] <- "metric"
  
  long_df <- raw %>%
    pivot_longer(cols = -metric, names_to = group_col_name, values_to = "value")
  
  median_df <- long_df %>%
    filter(metric == "median") %>%
    mutate(median = as.numeric(value)) %>%
    select(all_of(group_col_name), median)
  
  mean_df <- long_df %>%
    filter(metric == "mean") %>%
    mutate(mean = as.numeric(value)) %>%
    select(all_of(group_col_name), mean)
  
  hpd_df <- long_df %>%
    filter(metric == "95% HPD interval") %>%
    mutate(
      value     = as.character(value),
      hpd_lower = as.numeric(str_match(value, "\\[([^,]+),")[, 2]),
      hpd_upper = as.numeric(str_match(value, ",\\s*([^\\]]+)\\]")[, 2])
    ) %>%
    select(all_of(group_col_name), hpd_lower, hpd_upper)
  
  ess_df <- long_df %>%
    filter(metric == "effective sample size (ESS)") %>%
    mutate(ESS = as.numeric(value)) %>%
    select(all_of(group_col_name), ESS)
  
  median_df %>%
    left_join(mean_df, by = group_col_name) %>%
    left_join(hpd_df,  by = group_col_name) %>%
    left_join(ess_df,  by = group_col_name)
}

# Clean BEAST protein labels so plotting uses shared protein names
clean_protein_name <- function(x) {
  x %>%
    str_replace("^D[1-4]_", "") %>%
    str_replace("^Capsid$", "C") %>%
    str_replace("^Envelope$", "E")
}
```

# Protein-level analysis across DENV1 to DENV4

## Import and combine protein sheets

```{r protein-data}
protein_df <- bind_rows(
  parse_rate_sheet(file_path, "DENV1_Proteins", "protein_raw") %>% mutate(serotype = "DENV1"),
  parse_rate_sheet(file_path, "DENV2_Proteins", "protein_raw") %>% mutate(serotype = "DENV2"),
  parse_rate_sheet(file_path, "DENV3_Proteins", "protein_raw") %>% mutate(serotype = "DENV3"),
  parse_rate_sheet(file_path, "DENV4_Proteins", "protein_raw") %>% mutate(serotype = "DENV4")
) %>%
  mutate(
    protein = clean_protein_name(as.character(protein_raw)),
    serotype = factor(serotype, levels = c("DENV1", "DENV2", "DENV3", "DENV4")),
    protein = factor(protein, levels = c("C", "PrM", "E", "NS1", "NS2A", "NS2B", "NS3", "NS4A", "NS4B", "NS5"))
  ) %>%
  mutate(
    mean = suppressWarnings(as.numeric(mean)),
    median = suppressWarnings(as.numeric(median)),
    hpd_lower = suppressWarnings(as.numeric(hpd_lower)),
    hpd_upper = suppressWarnings(as.numeric(hpd_upper)),
    ESS = suppressWarnings(as.numeric(ESS))
  )

protein_df
```

## Keep plottable values only

This step removes rows where the mean or HPD values are missing or non-finite.

```{r protein-clean}
protein_plot_df <- protein_df %>%
  filter(
    !is.na(protein),
    is.finite(mean),
    is.finite(hpd_lower),
    is.finite(hpd_upper)
  )

protein_plot_df
```

## Quick summary table

```{r protein-summary}
protein_summary <- protein_plot_df %>%
  arrange(protein, serotype) %>%
  select(serotype, protein, mean, hpd_lower, hpd_upper, ESS)

knitr::kable(protein_summary, digits = 6)
```

## Plot mean and 95% HPD for all proteins across serotypes

By default, this uses a log10 y-axis starting at 1e-4 with intermediate breaks. The original x10^-4 axis is retained below as a commented option.

```{r}
# =========================================================
# Shared Y-axis for protein plots
# Default: log10 scale with intermediate ticks starting at 1e-4.
# Optional x10^-4 scale is included below and commented out.
# =========================================================
protein_y_min <- 1e-4
protein_y_max <- max(protein_plot_df$hpd_upper, na.rm = TRUE)

protein_y <- scale_y_log10(
  limits = c(protein_y_min, protein_y_max),
  breaks = rate_log_breaks,
  labels = rate_log_labels,
  expand = expansion(mult = c(0.02, 0.08))
)

protein_y_label <- expression(
  atop(
    bold("Evolutionary rate"),
    bold("(s/s/y, log scale)")
  )
)

# ---- Optional x10^-4 scale ----
# To use this instead, comment out the protein_y and protein_y_label above,
# then uncomment this block.
# protein_y_min_x10_4 <- floor(min(protein_plot_df$hpd_lower, na.rm = TRUE) * 1e4)
# protein_y_max_x10_4 <- ceiling(max(protein_plot_df$hpd_upper, na.rm = TRUE) * 1e4)
#
# protein_y <- scale_y_continuous(
#   limits = c(protein_y_min_x10_4 / 1e4, protein_y_max_x10_4 / 1e4),
#   breaks = seq(protein_y_min_x10_4, protein_y_max_x10_4, by = 1) / 1e4,
#   labels = function(x) x * 1e4,
#   expand = expansion(mult = c(0.02, 0.08))
# )
#
# protein_y_label <- expression("Evolutionary rate (" %*% 10^-4 * " subs/site/year)")
```




####Faceting by protein
```{r protein-by-protein-one-row, fig.width=18, fig.height=4.5}
# =========================================================
# 7) Protein-specific evolutionary rates by serotype
# =========================================================

serotype_labels <- c(
  "DENV1" = "DENV-1",
  "DENV2" = "DENV-2",
  "DENV3" = "DENV-3",
  "DENV4" = "DENV-4"
)

protein_plot_df <- protein_plot_df %>%
  mutate(
    protein = factor(
      protein,
      levels = c(
        "C", "PrM", "E", "NS1", "NS2A",
        "NS2B", "NS3", "NS4A", "NS4B", "NS5"
      ),
      labels = c(
        "C", "PrM", "E", "NS1", "NS2A",
        "NS2B", "NS3", "NS4A", "NS4B", "NS5"
      )
    ),
    serotype = factor(
      serotype,
      levels = c("DENV1", "DENV2", "DENV3", "DENV4")
    )
  )

pd_serotype <- position_dodge(width = 0.65)

p_serotype_by_protein <- ggplot(
  protein_plot_df,
  aes(
    x = protein,
    y = mean,
    color = serotype,
    fill = serotype
  )
) +
  geom_errorbar(
    aes(
      ymin = hpd_lower,
      ymax = hpd_upper
    ),
    color = "black",
    position = pd_serotype,
    width = 0.18,
    linewidth = 0.75
  ) +
  geom_point(
    position = pd_serotype,
    shape = 21,
    color = "black",
    stroke = 0.5,
    size = 4.5
  ) +
  scale_color_manual(
    values = serotype_colors,
    breaks = names(serotype_labels),
    labels = serotype_labels,
    guide = "none"
  ) +
  scale_fill_manual(
    values = serotype_colors,
    breaks = names(serotype_labels),
    labels = serotype_labels,
    name = "Serotype"
  ) +
  guides(
    fill = guide_legend(
      title = "Serotype",
      override.aes = list(
        shape = 21,
        color = "black",
        size = 4
      )
    )
  ) +
  protein_y +
  labs(
    x = "Protein",
    y = protein_y_label
  ) +
  theme_classic(base_size = 20) +
  theme(
    axis.title.x = element_text(
      size = 24,
      face = "bold"
    ),

    axis.title.y = element_text(
      size = 24,
      face = "bold"
    ),

    axis.text.x = element_text(
      size = 22,
    ),

    axis.text.y = element_text(
      size = 22,
    ),

    legend.position = "bottom",

    legend.title = element_text(
      size = 20,
      face = "bold"
    ),

    legend.text = element_text(
      size = 18
    )
  )

print(p_serotype_by_protein)
```


## Optional note on missing values

```{r protein-missing}
protein_missing <- protein_df %>%
  filter(!is.finite(mean) | !is.finite(hpd_lower) | !is.finite(hpd_upper)) %>%
  select(serotype, protein_raw, protein, mean, hpd_lower, hpd_upper, ESS)

knitr::kable(protein_missing, digits = 6)
```

# Original code you provided

The section below preserves your original serotype and genotype analysis code.

```{r serotype-genotype-data}
# =========================================================
# 1) Serotype-level data
# =========================================================
serotype_df <- parse_rate_sheet(file_path, "Serotype", "serotype") %>%
  mutate(
    serotype = case_when(
      str_detect(serotype, "1") ~ "DENV1",
      str_detect(serotype, "2") ~ "DENV2",
      str_detect(serotype, "3") ~ "DENV3",
      str_detect(serotype, "4") ~ "DENV4",
      TRUE ~ serotype
    ),
    serotype = factor(serotype, levels = c("DENV1", "DENV2", "DENV3", "DENV4"))
  )

# =========================================================
# 2) Genotype-level data
# =========================================================
genotype_df <- bind_rows(
  parse_rate_sheet(file_path, "DENV1_Genotypes", "genotype") %>% mutate(serotype = "DENV1"),
  parse_rate_sheet(file_path, "DENV2_Genotypes", "genotype") %>% mutate(serotype = "DENV2"),
  parse_rate_sheet(file_path, "DENV3_Genotypes", "genotype") %>% mutate(serotype = "DENV3"),
  parse_rate_sheet(file_path, "DENV4_Genotypes", "genotype") %>% mutate(serotype = "DENV4")
) %>%
  mutate(
    serotype = factor(serotype, levels = c("DENV1", "DENV2", "DENV3", "DENV4"))
  )

# Manual genotype order, left to right within each serotype facet
desired_order <- c(
  "1I", "1III", "1IV", "1V",
  "2I", "2II", "2III", "2V",
  "3I", "3II", "3III",
  "4I", "4II"
)
extra_levels <- setdiff(unique(as.character(genotype_df$genotype)), desired_order)
all_levels <- c(desired_order, sort(extra_levels))

genotype_df <- genotype_df %>%
  mutate(genotype = factor(as.character(genotype), levels = all_levels))

# =========================================================
# 3) Shared Y-axis limits for serotype and genotype rate plots
# =========================================================
y_min <- 3e-4
y_max <- 3e-3

shared_y <- scale_y_log10(
  limits = c(y_min, y_max),
  breaks = rate_log_breaks,
  labels = rate_log_labels,
  expand = expansion(mult = c(0.02, 0.08))
)

shared_y_label <- expression(
  atop(
    bold("Evolutionary rate"),
    bold("(s/s/y, log scale)")
  )
)
```

# Lineage metadata: full dataset versus subsampled dataset

This section compares the full 70% coverage genome metadata with the subsampled metadata used for downstream BEAST analyses. The plots are organized into three chunks:

1. Date-range and sample-size summaries by serotype and dataset.
2. Country contribution summaries and country-by-sample-size plots.
3. Continent-balance plots focused on Asia versus the Americas, shown as a distinct diverging percentage plot rather than another DENV-1 to DENV-4 stacked count plot.

## Convert sampled TSV files to CSV

Run this once if the sampled CSV files have not already been created.

```{r convert-sampled-tsv-to-csv, eval=FALSE}
sampled_tsv_files <- c(
  DENV1 = "~/Desktop/DENV_EVO_2025/Lineage_metadata/DENV1_sampled.tsv",
  DENV2 = "~/Desktop/DENV_EVO_2025/Lineage_metadata/DENV2_sampled.tsv",
  DENV3 = "~/Desktop/DENV_EVO_2025/Lineage_metadata/DENV3_sampled.tsv",
  DENV4 = "~/Desktop/DENV_EVO_2025/Lineage_metadata/DENV4_sampled.tsv"
)

for (serotype_name in names(sampled_tsv_files)) {
  sampled_df_tmp <- read.delim(
    sampled_tsv_files[[serotype_name]],
    fill = TRUE,
    quote = "",
    comment.char = ""
  )

  write.csv(
    sampled_df_tmp,
    file.path(
      "~/Desktop/DENV_EVO_2025/Lineage_metadata",
      paste0(serotype_name, "_sampled.csv")
    ),
    row.names = FALSE
  )
}
```

## Metadata helpers and import

```{r lineage-metadata-import, fig.width=13, fig.height=5}
# =========================================================
# Metadata file paths
# =========================================================
metadata_files <- tibble::tribble(
  ~serotype, ~dataset, ~file_path,
  "DENV1", "Full dataset",       "~/Desktop/DENV_EVO_2025/Lineage_metadata/DENV1_together.csv",
  "DENV2", "Full dataset",       "~/Desktop/DENV_EVO_2025/Lineage_metadata/DENV2_together.csv",
  "DENV3", "Full dataset",       "~/Desktop/DENV_EVO_2025/Lineage_metadata/DENV3_together.csv",
  "DENV4", "Full dataset",       "~/Desktop/DENV_EVO_2025/Lineage_metadata/DENV4_together.csv",
  "DENV1", "Subsampled dataset", "~/Desktop/DENV_EVO_2025/Lineage_metadata/DENV1_sampled.csv",
  "DENV2", "Subsampled dataset", "~/Desktop/DENV_EVO_2025/Lineage_metadata/DENV2_sampled.csv",
  "DENV3", "Subsampled dataset", "~/Desktop/DENV_EVO_2025/Lineage_metadata/DENV3_sampled.csv",
  "DENV4", "Subsampled dataset", "~/Desktop/DENV_EVO_2025/Lineage_metadata/DENV4_sampled.csv"
)

serotype_levels <- c("DENV1", "DENV2", "DENV3", "DENV4")
dataset_levels <- c("Full dataset", "Subsampled dataset")

# ---------------------------------------------------------
# Colors
# ---------------------------------------------------------
continent_colors <- c(
  "Asia"     = "#C2B709",
  "Americas" = "#2A332D",
  "Other"    = "grey70"
)

country_other_color <- "grey78"

# ---------------------------------------------------------
# Helpers for standardizing columns and dates
# ---------------------------------------------------------
find_first_existing_col <- function(df, candidates) {
  hit <- candidates[candidates %in% names(df)]
  if (length(hit) == 0) NA_character_ else hit[[1]]
}

standardize_continent <- function(x) {
  x <- as.character(x)
  case_when(
    str_detect(str_to_lower(x), "asia") ~ "Asia",
    str_detect(str_to_lower(x), "america|caribbean|latin") ~ "Americas",
    TRUE ~ "Other"
  )
}

standardize_region_from_country <- function(country, continent = NA_character_) {
  # Prefer a continent/region column when it exists, then fall back to country names.
  continent_std <- standardize_continent(continent)
  country_clean <- str_to_lower(str_squish(as.character(country)))

  asia_countries <- c(
    "bangladesh", "bhutan", "brunei", "cambodia", "china", "hong kong", "india",
    "indonesia", "japan", "laos", "malaysia", "maldives", "myanmar", "nepal",
    "pakistan", "philippines", "singapore", "south korea", "sri lanka", "taiwan",
    "thailand", "timor-leste", "vietnam"
  )

  americas_countries <- c(
    "anguilla", "antigua and barbuda", "argentina", "aruba", "bahamas", "barbados",
    "belize", "bolivia", "brazil", "canada", "cayman islands", "chile", "colombia",
    "costa rica", "cuba", "curacao", "dominica", "dominican republic", "ecuador",
    "el salvador", "french guiana", "grenada", "guadeloupe", "guatemala", "guyana",
    "haiti", "honduras", "jamaica", "martinique", "mexico", "nicaragua", "panama",
    "paraguay", "peru", "puerto rico", "saint barthelemy", "saint kitts and nevis",
    "saint lucia", "saint martin", "saint vincent and the grenadines", "suriname",
    "trinidad and tobago", "united states", "usa", "uruguay", "venezuela", "virgin islands"
  )

  case_when(
    continent_std %in% c("Asia", "Americas") ~ continent_std,
    country_clean %in% asia_countries ~ "Asia",
    country_clean %in% americas_countries ~ "Americas",
    str_detect(country_clean, "usa|united states|puerto rico|virgin islands") ~ "Americas",
    TRUE ~ "Other"
  )
}

extract_decimal_year <- function(x) {
  # Robust parser for years, decimal years, ISO dates, slash dates, and messy dates.
  # Falls back to the first 4-digit year if a full date cannot be parsed.
  if (inherits(x, "Date")) {
    yr <- as.numeric(format(x, "%Y"))
    doy <- as.numeric(format(x, "%j"))
    return(yr + (doy - 1) / 365.25)
  }

  x_chr <- str_trim(as.character(x))
  x_chr[x_chr %in% c("", "NA", "N/A", "na", "unknown", "Unknown", "?")] <- NA_character_

  x_num <- suppressWarnings(as.numeric(x_chr))
  out <- rep(NA_real_, length(x_chr))

  decimal_or_plain_year <- !is.na(x_num) & x_num >= 1900 & x_num <= 2100
  out[decimal_or_plain_year] <- x_num[decimal_or_plain_year]

  excel_serial <- !is.na(x_num) & x_num > 20000 & x_num < 80000
  if (any(excel_serial)) {
    d <- as.Date(x_num[excel_serial], origin = "1899-12-30")
    yr <- as.numeric(format(d, "%Y"))
    doy <- as.numeric(format(d, "%j"))
    out[excel_serial] <- yr + (doy - 1) / 365.25
  }

  still_missing <- is.na(out) & !is.na(x_chr)
  date_formats <- c(
    "%Y-%m-%d", "%Y/%m/%d", "%Y.%m.%d",
    "%m/%d/%Y", "%m-%d-%Y", "%m.%d.%Y",
    "%d/%m/%Y", "%d-%m-%Y", "%d.%m.%Y",
    "%Y-%m", "%Y/%m", "%Y.%m"
  )

  for (fmt in date_formats) {
    idx <- which(still_missing)
    if (length(idx) == 0) break
    d <- suppressWarnings(as.Date(x_chr[idx], format = fmt))
    good <- !is.na(d)
    if (any(good)) {
      yr <- as.numeric(format(d[good], "%Y"))
      doy <- as.numeric(format(d[good], "%j"))
      out[idx[good]] <- yr + (doy - 1) / 365.25
    }
    still_missing <- is.na(out) & !is.na(x_chr)
  }

  still_missing <- is.na(out) & !is.na(x_chr)
  fallback_year <- suppressWarnings(as.numeric(str_extract(x_chr[still_missing], "(19|20)\\d{2}")))
  out[still_missing] <- fallback_year

  out
}

read_metadata_file <- function(file_path, serotype_label, dataset_label) {
  df <- read_csv(file_path, show_col_types = FALSE)

  country_col <- find_first_existing_col(df, c("country", "Country", "COUNTRY", "location_country", "Location_Country"))
  continent_col <- find_first_existing_col(df, c("Continent", "continent", "CONTINENT", "region", "Region"))
  lineage_col <- find_first_existing_col(df, c("major_lineage", "Major_Lineage", "Major lineage", "lineage", "Lineage"))
  genotype_col <- find_first_existing_col(df, c("genotype", "Genotype", "GENOTYPE"))
  date_col <- find_first_existing_col(df, c(
    "date", "Date", "collection_date", "Collection_Date", "Collection date",
    "sampling_date", "Sampling_Date", "sample_date", "Sample_Date", "year", "Year"
  ))

  if (is.na(country_col)) {
    df <- df %>% mutate(country = "Unknown")
  } else {
    df <- df %>% mutate(country = as.character(.data[[country_col]]))
  }

  if (is.na(continent_col)) {
    df <- df %>% mutate(Continent = "Other")
  } else {
    df <- df %>% mutate(Continent = .data[[continent_col]])
  }

  if (is.na(lineage_col)) {
    df <- df %>% mutate(major_lineage = NA_character_)
  } else {
    df <- df %>% mutate(major_lineage = as.character(.data[[lineage_col]]))
  }

  if (is.na(genotype_col)) {
    df <- df %>% mutate(genotype = NA_character_)
  } else {
    df <- df %>% mutate(genotype = as.character(.data[[genotype_col]]))
  }

  if (is.na(date_col)) {
    stop(paste0("No date/year column found in ", file_path))
  }

  df %>%
    mutate(
      serotype = serotype_label,
      dataset = dataset_label,
      country = if_else(is.na(country) | country == "", "Unknown", country),
      Continent = factor(standardize_region_from_country(country, Continent), levels = c("Asia", "Americas", "Other")),
      sample_year = extract_decimal_year(.data[[date_col]]),
      year = floor(sample_year)
    ) %>%
    select(dataset, serotype, genotype, major_lineage, country, Continent, sample_year, year, everything())
}

metadata_df <- purrr::pmap_dfr(
  metadata_files,
  function(serotype, dataset, file_path) {
    read_metadata_file(file_path, serotype, dataset)
  }
) %>%
  mutate(
    dataset = factor(dataset, levels = dataset_levels),
    serotype = factor(serotype, levels = serotype_levels)
  )

# Keep these names for downstream chunks that use the subsampled metadata.
sampled_df <- metadata_df %>% filter(dataset == "Subsampled dataset")
sampled_df_raw <- sampled_df
```

## Serotype and genotype evolutionary rates with flipped sample-size panels

This figure places the BEAST evolutionary-rate estimates in the top row and the genome sample counts directly below the matching serotype and genotype panels. The sample-size y-axis is reversed so that bars extend downward from zero.

```{r serotype-genotype-rates-with-sample-size, fig.width=15, fig.height=6.5}
# =========================================================
# 1) Sample-size data from genomes INCLUDED in rate analyses
# =========================================================

serotype_n_df <- serotype_df %>%
  mutate(
    serotype = as.character(serotype)
  ) %>%
  left_join(
    sampled_df %>%
      filter(!is.na(serotype)) %>%
      count(serotype, name = "n"),
    by = "serotype"
  ) %>%
  mutate(
    serotype = factor(
      serotype,
      levels = c("DENV1", "DENV2", "DENV3", "DENV4")
    )
  )

genotype_n_df <- genotype_df %>%
  mutate(
    genotype = as.character(genotype)
  ) %>%
  left_join(
    sampled_df %>%
      filter(
        !is.na(serotype),
        !is.na(genotype),
        genotype != ""
      ) %>%
      count(serotype, genotype, name = "n"),
    by = c("serotype", "genotype")
  ) %>%
  mutate(
    serotype = factor(
      serotype,
      levels = c("DENV1", "DENV2", "DENV3", "DENV4")
    ),
    genotype = factor(
      genotype,
      levels = all_levels
    )
  )

# =========================================================
# 2) Shared flipped y-axis for sample-size panels
# =========================================================

sample_y_max <- ceiling(
  max(
    c(serotype_n_df$n, genotype_n_df$n),
    na.rm = TRUE
  ) / 100
) * 100

sample_y_flipped <- scale_y_continuous(
  limits = c(-sample_y_max, 0),
  breaks = -seq(0, sample_y_max, by = 250),
  labels = function(x) ifelse(abs(x) %% 500 == 0, abs(x), ""),
  expand = expansion(mult = c(0.02, 0.02))
)

# =========================================================
# 3) SEROTYPE RATE PANEL
# =========================================================

p_serotype_rate <- ggplot(
  serotype_df,
  aes(
    x = serotype,
    y = mean,
    color = serotype
  )
) +
  scale_x_discrete(
  labels = c(
    "DENV1" = "DENV-1",
    "DENV2" = "DENV-2",
    "DENV3" = "DENV-3",
    "DENV4" = "DENV-4"
  )
) +
  geom_errorbar(
    aes(
      ymin = hpd_lower,
      ymax = hpd_upper
    ),
    color = "black",
    width = 0.18,
    linewidth = 0.8
  ) +
  geom_point(
    aes(fill = serotype),
    shape = 21,
    color = "black",
    stroke = 0.5,
    size = 3
  ) +
  scale_color_manual(
    values = serotype_colors,
    guide = "none"
  ) +
  scale_fill_manual(
  values = serotype_colors,
  guide = "none"
) +
scale_x_discrete(
  labels = c(
    "DENV1" = "DENV-1",
    "DENV2" = "DENV-2",
    "DENV3" = "DENV-3",
    "DENV4" = "DENV-4"
  )
) +
shared_y +
  labs(
    x = "Serotype",
    y = shared_y_label
  ) +
  theme_classic(base_size = 18) +
  theme(
    plot.title = element_text(size = 18, face = "bold"),

    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),

    axis.text.x = element_text(size = 16),
    axis.text.y = element_text(size = 18),

    axis.ticks.x = element_line(linewidth = 0.4),
    axis.ticks.y = element_line(linewidth = 0.4),

    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14, face = "bold")
  )

# =========================================================
# 4) GENOTYPE RATE PANEL
# =========================================================

genotype_serotype_label_df <- genotype_df %>%
  mutate(
    serotype = factor(
      serotype,
      levels = c("DENV1", "DENV2", "DENV3", "DENV4")
    ),
    genotype = factor(
      genotype,
      levels = all_levels
    )
  ) %>%
  filter(!is.na(genotype)) %>%
  group_by(serotype) %>%
  summarise(
    genotype = genotype[ceiling(n_distinct(genotype) / 2)],
    .groups = "drop"
  ) %>%
  mutate(
    serotype_label = recode(
      as.character(serotype),
      "DENV1" = "DENV-1",
      "DENV2" = "DENV-2",
      "DENV3" = "DENV-3",
      "DENV4" = "DENV-4"
    )
  )

p_genotype_rate <- ggplot(
  genotype_df,
  aes(
    x = genotype,
    y = mean,
    color = serotype
  )
) +
  geom_errorbar(
    aes(
      ymin = hpd_lower,
      ymax = hpd_upper
    ),
    color = "black",
    width = 0.18,
    linewidth = 0.7
  ) +
  geom_point(
    aes(fill = serotype),
    shape = 21,
    color = "black",
    stroke = 0.5,
    size = 2.6
  ) +
  geom_label(
    data = genotype_serotype_label_df,
    aes(
      x = genotype,
      y = Inf,
      label = serotype_label
    ),
    inherit.aes = FALSE,
    vjust = -0.5,
    fontface = "bold",
    size = 7,
    label.size = 0.35,
    label.padding = unit(0.16, "lines"),
    fill = "white"
  ) +
  scale_color_manual(
    values = serotype_colors,
    guide = "none"
  ) +
  scale_fill_manual(
    values = serotype_colors,
    guide = "none"
  ) +
  facet_grid(
    cols = vars(serotype),
    scales = "free_x",
    space = "free_x"
  ) +
  shared_y +
  coord_cartesian(clip = "off") +
  labs(
    x = "Genotype",
    y = NULL
  ) +
  theme_classic(base_size = 18) +
  theme(
    plot.title = element_text(
      size = 18,
      face = "bold",
      hjust = 0.5
    ),

    strip.text.x = element_blank(),

    strip.background = element_blank(),

    axis.title.x = element_text(
      size = 18,
      face = "bold"
    ),

    axis.title.y = element_text(
      size = 18,
      face = "bold"
    ),

    axis.text.x = element_text(
      size = 18
    ),

    axis.text.y = element_blank(),

    axis.ticks.x = element_line(
      linewidth = 0.4
    ),

    axis.ticks.y = element_blank(),

    panel.spacing.x = unit(0.15, "lines"),

    plot.margin = margin(34, 5, 0, 0)
  )

# =========================================================
# 5) SEROTYPE SAMPLE-SIZE PANEL
# =========================================================

p_serotype_samples <- ggplot(
  serotype_n_df,
  aes(
    x = serotype,
    y = -n
  )
) +
  geom_col(
    width = 0.68,
    fill = "grey70",
    color = "black"
  ) +
  scale_x_discrete(position = "top") +
  sample_y_flipped +
  labs(
    x = NULL,
    y = "N"
  ) +
  theme_classic(base_size = 18) +
  theme(
  axis.text.x = element_blank(),

  axis.ticks.x = element_blank(),

  axis.line.x.top = element_line(
    color = "black",
    linewidth = 0.4
  ),

  axis.ticks.x.top = element_line(
    color = "black",
    linewidth = 0.4
  ),

  axis.title.x = element_text(
    size = 18,
    face = "bold"
  ),

  axis.title.y = element_text(
    size = 18,
    face = "bold"
  ),

  axis.text.y = element_text(
    size = 18
  ),

  axis.ticks.y = element_line(
    linewidth = 0.4
  ),

  plot.title = element_text(
    size = 16,
    face = "bold",
    hjust = 0.5
  ),

  plot.margin = margin(0, 10, 5, 5)
)

# =========================================================
# 6) GENOTYPE SAMPLE-SIZE PANEL
# =========================================================

p_genotype_samples <- ggplot(
  genotype_n_df,
  aes(
    x = genotype,
    y = -n
  )
) +
  geom_col(
    width = 0.68,
    fill = "grey70",
    color = "black"
  ) +
  facet_grid(
    cols = vars(serotype),
    scales = "free_x",
    space = "free_x"
  ) +
  scale_x_discrete(position = "top") +
  sample_y_flipped +
  labs(
    x = NULL,
    y = NULL
  ) +
  theme_classic(base_size = 16) +
  theme(
  strip.text.x = element_blank(),

  strip.background = element_blank(),

  axis.title.x = element_text(
    size = 16,
    face = "bold"
  ),

  axis.title.y = element_text(
    size = 16,
    face = "bold"
  ),

  axis.text.x = element_blank(),

  axis.text.y = element_blank(),

  axis.ticks.x = element_blank(),

  axis.ticks.y = element_blank(),

  axis.line.x.top = element_line(
    color = "black",
    linewidth = 0.4
  ),

  axis.ticks.x.top = element_line(
    color = "black",
    linewidth = 0.4
  ),

  plot.title = element_text(
    size = 16,
    face = "bold",
    hjust = 0.5
  ),

  panel.spacing.x = unit(0.15, "lines"),

  plot.margin = margin(0, 5, 5, 0)
)

# =========================================================
# 7) BUILD ROWS FIRST
# =========================================================

p_rate_row <-
  p_serotype_rate +
  p_genotype_rate +
  plot_layout(widths = c(1.5, 3.5))

p_sample_row <-
  p_serotype_samples +
  p_genotype_samples +
  plot_layout(widths = c(1.5, 3.5))

# =========================================================
# 8) FINAL COMBINED FIGURE
# =========================================================

p_serotype_genotype_with_samples <-
  p_rate_row /
  p_sample_row +
  plot_layout(
    heights = c(2.4, 1)
  )

print(p_serotype_genotype_with_samples)
```


## Date range and sample size summaries

```{r metadata-date-range-summary}
date_range_summary <- metadata_df %>%
  group_by(dataset, serotype) %>%
  summarise(
    n_sequences = n(),
    first_year = floor(min(sample_year, na.rm = TRUE)),
    last_year = floor(max(sample_year, na.rm = TRUE)),
    date_range = paste0(first_year, "-", last_year),
    n_countries = n_distinct(country),
    .groups = "drop"
  ) %>%
  arrange(dataset, serotype)

knitr::kable(date_range_summary)
```

```{r metadata-sample-size-overview, fig.width=10, fig.height=4.5}
metadata_sample_size_summary <- metadata_df %>%
  count(dataset, serotype, name = "n_sequences")

p_metadata_sample_size <- ggplot(
  metadata_sample_size_summary,
  aes(x = serotype, y = n_sequences, fill = dataset)
) +
  geom_col(position = position_dodge(width = 0.75), color = "black", width = 0.68) +
  geom_text(
    aes(label = comma(n_sequences)),
    position = position_dodge(width = 0.75),
    vjust = -0.25,
    size = 3.2
  ) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Full versus subsampled metadata size by serotype",
    x = "Serotype",
    y = "Number of sequences",
    fill = "Dataset"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom"
  )

p_metadata_sample_size
```

## Country contribution summaries and country-by-sample-size plots

```{r country-contribution-summaries}
country_counts <- metadata_df %>%
  filter(!is.na(country), country != "", country != "Unknown") %>%
  count(dataset, serotype, country, name = "n") %>%
  group_by(dataset, serotype) %>%
  mutate(percent = 100 * n / sum(n)) %>%
  ungroup()

country_extremes_by_serotype <- country_counts %>%
  group_by(dataset, serotype) %>%
  summarise(
    most_contributing_country = country[which.max(n)],
    most_contributing_n = max(n),
    most_contributing_percent = percent[which.max(n)],
    least_contributing_country = country[which.min(n)],
    least_contributing_n = min(n),
    least_contributing_percent = percent[which.min(n)],
    .groups = "drop"
  ) %>%
  arrange(dataset, serotype)

country_extremes_overall <- metadata_df %>%
  filter(!is.na(country), country != "", country != "Unknown") %>%
  count(dataset, country, name = "n") %>%
  group_by(dataset) %>%
  mutate(percent = 100 * n / sum(n)) %>%
  summarise(
    most_contributing_country = country[which.max(n)],
    most_contributing_n = max(n),
    most_contributing_percent = percent[which.max(n)],
    least_contributing_country = country[which.min(n)],
    least_contributing_n = min(n),
    least_contributing_percent = percent[which.min(n)],
    .groups = "drop"
  ) %>%
  arrange(dataset)

knitr::kable(country_extremes_overall, digits = 2)
knitr::kable(country_extremes_by_serotype, digits = 2)
```

```{r country-sample-size-plots, fig.width=13, fig.height=9}
# Collapse small countries within each dataset and serotype so the country plots remain readable.
country_min_n <- 5
country_plot_df <- country_counts %>%
  mutate(
    country_plot = if_else(n >= country_min_n, country, "Other countries")
  ) %>%
  group_by(dataset, serotype, country_plot) %>%
  summarise(n = sum(n), .groups = "drop") %>%
  group_by(dataset, serotype) %>%
  mutate(percent = 100 * n / sum(n)) %>%
  ungroup()

country_levels <- country_plot_df %>%
  group_by(country_plot) %>%
  summarise(total_n = sum(n), .groups = "drop") %>%
  arrange(desc(total_n)) %>%
  pull(country_plot)

country_levels <- c(setdiff(country_levels, "Other countries"), "Other countries")

country_colors <- c(
  setNames(
    colorspace::qualitative_hcl(length(setdiff(country_levels, "Other countries")), palette = "Dynamic"),
    setdiff(country_levels, "Other countries")
  ),
  "Other countries" = country_other_color
)

if (nrow(country_plot_df) == 0) {
  message(
    "No country rows were available for the country sample-size plot. Check country values with: count(metadata_df, dataset, serotype, country, sort = TRUE)"
  )
} else {
  p_country_sample_size <- ggplot(
    country_plot_df %>% mutate(country_plot = factor(country_plot, levels = country_levels)),
    aes(x = serotype, y = n, fill = country_plot)
  ) +
    geom_col(color = "black", linewidth = 0.25) +
    facet_wrap(~ dataset, ncol = 1, scales = "free_y", drop = FALSE) +
    scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.06))) +
    scale_fill_manual(values = country_colors, drop = FALSE) +
    labs(
      title = paste0("Country contribution by sample size, countries with n < ", country_min_n, " collapsed"),
      x = "Serotype",
      y = "Number of sequences",
      fill = "Country"
    ) +
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      strip.text = element_text(face = "bold"),
      axis.title = element_text(face = "bold"),
      axis.text.x = element_text(face = "bold"),
      legend.position = "bottom",
      legend.text = element_text(size = 8),
      legend.key.size = unit(0.35, "cm")
    ) +
    guides(fill = guide_legend(ncol = 6, byrow = TRUE))

  p_country_sample_size
}
```

## Asia, Americas, and other regions as a share of total sequences

This plot uses additive 100% stacked bars. Each bar is calculated from the full serotype-specific dataset total, so Asia and the Americas are interpreted relative to all sequences, with `Other` retained in the denominator and shown directly in the bar.

```{r asia-americas-balance-plot, fig.width=11, fig.height=5.5}
continent_balance_df <- metadata_df %>%
  mutate(
    dataset = factor(as.character(dataset), levels = dataset_levels),
    serotype = factor(as.character(serotype), levels = serotype_levels),
    Continent = factor(as.character(Continent), levels = c("Asia", "Americas", "Other"))
  ) %>%
  count(dataset, serotype, Continent, name = "n", .drop = FALSE) %>%
  tidyr::complete(
    dataset = factor(dataset_levels, levels = dataset_levels),
    serotype = factor(serotype_levels, levels = serotype_levels),
    Continent = factor(c("Asia", "Americas", "Other"), levels = c("Asia", "Americas", "Other")),
    fill = list(n = 0)
  ) %>%
  group_by(dataset, serotype) %>%
  mutate(
    dataset_serotype_total = sum(n, na.rm = TRUE),
    percent = if_else(dataset_serotype_total > 0, 100 * n / dataset_serotype_total, 0)
  ) %>%
  ungroup() %>%
  filter(dataset_serotype_total > 0) %>%
  mutate(
    Continent = factor(Continent, levels = c("Other", "Americas", "Asia")),
    percent_label = if_else(percent >= 4, paste0(round(percent, 1), "%"), "")
  )

if (nrow(continent_balance_df) == 0) {
  message(
    "No rows were available for the region contribution plot. Check metadata_df with: count(metadata_df, dataset, serotype, Continent, sort = TRUE)"
  )
} else {
  p_asia_americas_balance <- ggplot(
    continent_balance_df,
    aes(x = serotype, y = percent, fill = Continent)
  ) +
    geom_col(width = 0.72, color = "black", linewidth = 0.25) +
    geom_text(
      aes(label = percent_label),
      position = position_stack(vjust = 0.5),
      size = 3.2,
      color = "black"
    ) +
    facet_wrap(~ dataset, nrow = 1, drop = FALSE) +
    scale_fill_manual(values = continent_colors[c("Other", "Americas", "Asia")], drop = FALSE) +
    scale_y_continuous(
      limits = c(0, 100),
      breaks = seq(0, 100, by = 25),
      labels = function(x) paste0(x, "%"),
      expand = expansion(mult = c(0, 0.02))
    ) +
    labs(
      title = "Regional contribution in full and subsampled datasets",
      subtitle = "Each bar sums to 100% of all sequences for that serotype and dataset",
      x = "Serotype",
      y = "Share of total sequences",
      fill = "Region"
    ) +
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5),
      strip.text = element_text(face = "bold"),
      axis.title = element_text(face = "bold"),
      axis.text.x = element_text(face = "bold"),
      legend.position = "bottom",
      legend.title = element_text(face = "bold")
    )

  p_asia_americas_balance
}
```

## Subsampled dataset lineage plots by serotype

The lineage plots are still arranged vertically because the legends are smaller. Country-by-year plots are printed one serotype at a time below so the x-axis and legends remain readable.

```{r subsampled-lineage-and-country-plots, fig.width=13, fig.height=8}
plot_year_by_lineage <- function(df, serotype_label) {
  year_summary <- df %>%
    filter(serotype == serotype_label, !is.na(year), !is.na(major_lineage), major_lineage != "") %>%
    count(year, major_lineage, name = "n")

  lineage_cols <- setNames(
    colorspace::qualitative_hcl(length(unique(year_summary$major_lineage)), palette = "Dark3"),
    sort(unique(year_summary$major_lineage))
  )

  ggplot(year_summary, aes(x = year, y = n, fill = major_lineage)) +
    geom_col() +
    scale_fill_manual(values = lineage_cols) +
    scale_x_continuous(breaks = sort(unique(year_summary$year))) +
    labs(
      title = paste0(str_replace(serotype_label, "DENV", "DENV-"), " subsampled sequences by calendar year"),
      x = "Calendar year",
      y = "Count",
      fill = "Major lineage"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "bottom",
      legend.text = element_text(size = 9),
      legend.key.size = unit(0.35, "cm")
    ) +
    guides(fill = guide_legend(ncol = 9, byrow = TRUE))
}

plot_year_by_country <- function(df, serotype_label, dataset_label, top_n_countries = 12) {
  serotype_df <- df %>%
    filter(serotype == serotype_label, !is.na(year), !is.na(country), country != "", country != "Unknown")

  top_countries <- serotype_df %>%
    count(country, sort = TRUE) %>%
    slice_head(n = top_n_countries) %>%
    pull(country)

  year_summary <- serotype_df %>%
    mutate(country_plot = if_else(country %in% top_countries, country, "Other countries")) %>%
    count(year, country_plot, name = "n")

  country_levels_local <- year_summary %>%
    group_by(country_plot) %>%
    summarise(total_n = sum(n), .groups = "drop") %>%
    arrange(desc(total_n)) %>%
    pull(country_plot)
  country_levels_local <- c(setdiff(country_levels_local, "Other countries"), "Other countries")

  cols <- c(
    setNames(
      colorspace::qualitative_hcl(length(setdiff(country_levels_local, "Other countries")), palette = "Dynamic"),
      setdiff(country_levels_local, "Other countries")
    ),
    "Other countries" = country_other_color
  )

  ggplot(year_summary %>% mutate(country_plot = factor(country_plot, levels = country_levels_local)),
         aes(x = year, y = n, fill = country_plot)) +
    geom_col(width = 0.95) +
    scale_fill_manual(values = cols, drop = FALSE) +
    scale_x_continuous(
      breaks = scales::pretty_breaks(n = 12),
      expand = expansion(mult = c(0.005, 0.01))
    ) +
    scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.06))) +
    labs(
      title = paste0(str_replace(serotype_label, "DENV", "DENV-"), " ", str_to_lower(dataset_label), " sequences by country and calendar year"),
      subtitle = paste0("Top ", top_n_countries, " countries are shown individually; remaining countries are collapsed"),
      x = "Calendar year",
      y = "Count",
      fill = "Country"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "bottom",
      legend.text = element_text(size = 8),
      legend.key.size = unit(0.32, "cm")
    ) +
    guides(fill = guide_legend(ncol = 5, byrow = TRUE))
}

subsampled_lineage_plots <- purrr::map(serotype_levels, ~ plot_year_by_lineage(sampled_df, .x))
print(wrap_plots(subsampled_lineage_plots, ncol = 1))
```

## Subsampled dataset country-by-year plots by serotype

```{r subsampled-country-year-plots-readable, fig.width=13, fig.height=5.6, results='asis'}
for (serotype_i in serotype_levels) {
  print(plot_year_by_country(sampled_df, serotype_i, "Subsampled dataset", top_n_countries = 12))
}
```

## Full dataset lineage plots by serotype

```{r full-lineage-plots, fig.width=13, fig.height=8}
full_df <- metadata_df %>% filter(dataset == "Full dataset")

full_lineage_plots <- purrr::map(serotype_levels, ~ plot_year_by_lineage(full_df, .x))
print(wrap_plots(full_lineage_plots, ncol = 1))
```

## Full dataset country-by-year plots by serotype

```{r full-country-year-plots-readable, fig.width=13, fig.height=5.6, results='asis'}
for (serotype_i in serotype_levels) {
  print(plot_year_by_country(full_df, serotype_i, "Full dataset", top_n_countries = 12))
}
```

```{r filtered-subsampled-objects-for-major-lineage-rate-plot}
# ---------------------------------------------------------
# Keep only target genotypes for downstream major-lineage plots
# ---------------------------------------------------------
desired_genotypes <- c(
  "1I", "1III", "1IV", "1V",
  "2I", "2II", "2III", "2V",
  "3I", "3II", "3III",
  "4I", "4II"
)

filtered_sampled_df <- sampled_df %>%
  filter(genotype %in% desired_genotypes) %>%
  mutate(
    genotype = factor(genotype, levels = desired_genotypes),
    serotype = factor(as.character(serotype), levels = serotype_levels)
  )
```

```{r major-lineage-aligned-evolutionary-rates-sampling, fig.width=18, fig.height=7.5}
# =========================================================
# Major lineage evolutionary rates with aligned sample size,
# continent distribution, and temporal sampling evidence.
#
# Alignment fix:
# - Use one master lineage order.
# - Convert lineage identity to one numeric x_index used by all panels.
# - Do not facet with free_x for the aligned figure.
# - Keep identical x limits in all panels.
# =========================================================

min_major_lineage_n <- 20
collection_point_size <- 0.55
collection_linewidth <- 0.45

# Parse the rate estimates first, then use them as the master set of plotted lineages.
major_lineage_df <- bind_rows(
  parse_rate_sheet(file_path, "DENV1_Major_Lineages", "major_lineage") %>% mutate(serotype = "DENV1"),
  parse_rate_sheet(file_path, "DENV2_Major_Lineages", "major_lineage") %>% mutate(serotype = "DENV2"),
  parse_rate_sheet(file_path, "DENV3_Major_Lineages", "major_lineage") %>% mutate(serotype = "DENV3"),
  parse_rate_sheet(file_path, "DENV4_Major_Lineages", "major_lineage") %>% mutate(serotype = "DENV4")
) %>%
  mutate(
    serotype = factor(serotype, levels = c("DENV1", "DENV2", "DENV3", "DENV4")),
    major_lineage = as.character(major_lineage)
  ) %>%
  filter(
    str_detect(major_lineage, "_"),
    is.finite(mean),
    is.finite(hpd_lower),
    is.finite(hpd_upper)
  )

valid_major_lineages <- major_lineage_df %>%
  distinct(serotype, major_lineage)

lineage_n_df <- filtered_sampled_df %>%
  filter(!is.na(major_lineage), major_lineage != "", str_detect(major_lineage, "_")) %>%
  mutate(
    serotype = factor(as.character(serotype), levels = c("DENV1", "DENV2", "DENV3", "DENV4")),
    genotype = factor(as.character(genotype), levels = desired_genotypes),
    Continent = factor(as.character(Continent), levels = c("Asia", "Americas", "Other"))
  ) %>%
  semi_join(valid_major_lineages, by = c("serotype", "major_lineage")) %>%
  count(serotype, genotype, major_lineage, Continent, name = "n")

lineage_totals <- lineage_n_df %>%
  group_by(serotype, genotype, major_lineage) %>%
  summarise(n_total = sum(n), .groups = "drop") %>%
  filter(n_total >= min_major_lineage_n) %>%
  mutate(
    serotype = factor(as.character(serotype), levels = c("DENV1", "DENV2", "DENV3", "DENV4")),
    genotype = factor(as.character(genotype), levels = desired_genotypes),
    lineage_suffix = str_remove(major_lineage, paste0("^", as.character(genotype), "_?")),
    lineage_suffix = if_else(lineage_suffix == major_lineage, "", lineage_suffix),
    lineage_suffix_sort = if_else(lineage_suffix == "", "0", lineage_suffix)
  ) %>%
  arrange(serotype, genotype, lineage_suffix_sort, major_lineage)

lineage_order <- lineage_totals$major_lineage
lineage_index_df <- lineage_totals %>%
  mutate(
    x_index = row_number(),
    lineage_label = paste0(major_lineage),
    serotype_label = recode(as.character(serotype),
      "DENV1" = "DENV-1",
      "DENV2" = "DENV-2",
      "DENV3" = "DENV-3",
      "DENV4" = "DENV-4"
    ),
    genotype_label = as.character(genotype)
  ) %>%
  select(serotype, genotype, major_lineage, n_total, x_index, lineage_label, serotype_label, genotype_label)

x_breaks <- lineage_index_df$x_index
x_labels <- lineage_index_df$lineage_label
x_limits <- c(0.4, max(x_breaks) + 0.6)

genotype_label_df <- lineage_index_df %>%
  group_by(serotype, genotype, serotype_label, genotype_label) %>%
  summarise(
    xmin = min(x_index) - 0.5,
    xmax = max(x_index) + 0.5,
    x_mid = mean(range(x_index)),
    .groups = "drop"
  )

serotype_label_df <- lineage_index_df %>%
  group_by(serotype, serotype_label) %>%
  summarise(
    xmin = min(x_index) - 0.5,
    xmax = max(x_index) + 0.5,
    x_mid = mean(range(x_index)),
    .groups = "drop"
  )

serotype_boundary_df <- serotype_label_df %>%
  filter(xmin > min(xmin, na.rm = TRUE))

serotype_box_df <- serotype_label_df %>%
  mutate(y = Inf)

major_lineage_df <- major_lineage_df %>%
  inner_join(lineage_index_df, by = c("serotype", "major_lineage"))

major_lineage_y_min <- 3e-4
major_lineage_y_max <- max(major_lineage_df$hpd_upper, na.rm = TRUE)

major_lineage_y <- scale_y_log10(
  limits = c(major_lineage_y_min, major_lineage_y_max),
  breaks = rate_log_breaks,
  labels = rate_log_labels,
  expand = expansion(mult = c(0.02, 0.12))
)

major_lineage_y_label <- expression(
  atop(
    bold("Evolutionary rate"),
    bold("(s/s/y, log scale)")
  )
)

lineage_bar_df <- lineage_n_df %>%
  inner_join(lineage_index_df, by = c("serotype", "genotype", "major_lineage")) %>%
  select(x_index, Continent, n) %>%
  complete(
    x_index = x_breaks,
    Continent = factor(c("Asia", "Americas", "Other"), levels = c("Asia", "Americas", "Other")),
    fill = list(n = 0)
  )

lineage_bar_totals <- lineage_index_df %>%
  transmute(x_index, n_total)

sample_y_max <- ceiling(max(lineage_index_df$n_total, na.rm = TRUE) / 50) * 50

lineage_presence_years <- filtered_sampled_df %>%
  filter(!is.na(major_lineage), major_lineage != "", str_detect(major_lineage, "_"), is.finite(sample_year)) %>%
  mutate(
    year = floor(sample_year),
    serotype = factor(as.character(serotype), levels = c("DENV1", "DENV2", "DENV3", "DENV4")),
    genotype = factor(as.character(genotype), levels = desired_genotypes),
    Continent = factor(as.character(Continent), levels = c("Asia", "Americas", "Other"))
  ) %>%
  distinct(serotype, genotype, major_lineage, Continent, year) %>%
  inner_join(lineage_index_df, by = c("serotype", "genotype", "major_lineage")) %>%
  arrange(x_index, Continent, year) %>%
  group_by(x_index, Continent) %>%
  mutate(run_id = cumsum(year - lag(year, default = first(year)) > 1)) %>%
  ungroup()

full_start_year <- floor(min(lineage_presence_years$year, na.rm = TRUE))
full_end_year <- ceiling(max(lineage_presence_years$year, na.rm = TRUE))

temporal_runs_table <- lineage_presence_years %>%
  group_by(serotype, genotype, major_lineage, n_total) %>%
  summarise(
    first_year = min(year),
    last_year = max(year),
    years_sampled = n_distinct(year),
    .groups = "drop"
  ) %>%
  arrange(serotype, genotype, major_lineage)

knitr::kable(temporal_runs_table)

shared_x <- scale_x_continuous(
  limits = x_limits,
  breaks = x_breaks,
  labels = x_labels,
  expand = expansion(mult = c(0, 0))
)

shared_plot_theme <- theme_classic(base_size = 20) +
  theme(
  plot.title = element_text(
    size = 20,
    face = "bold",
    hjust = 0.5
  ),

  axis.title.x = element_blank(),

  axis.title.y = element_text(
    size = 18,
    face = "bold"
  ),

  axis.text.x = element_blank(),

  axis.text.y = element_text(
    size = 18
  ),

  axis.ticks.x = element_blank(),

  axis.ticks.y = element_line(
    linewidth = 0.4
  ),

  legend.title = element_text(
    size = 18,
    face = "bold"
  ),

  legend.text = element_text(
    size = 16
  ),

  legend.position = "none",

  plot.margin = margin(8, 8, 8, 8)
)

p_major_lineage_rate_aligned <- ggplot(
  major_lineage_df,
  aes(x = x_index, y = mean)
) +
  geom_errorbar(
    aes(ymin = hpd_lower, ymax = hpd_upper),
    width = 0.14,
    linewidth = 0.65,
    color = "black"
  ) +
  geom_point(
    aes(fill = serotype),
    shape = 21,
    color = "black",
    stroke = 0.45,
    size = 2.6
  ) +
  geom_vline(
    data = genotype_label_df,
    aes(xintercept = xmin),
    inherit.aes = FALSE,
    color = "grey82",
    linewidth = 0.35
  ) +
  geom_vline(
    data = serotype_boundary_df,
    aes(xintercept = xmin),
    inherit.aes = FALSE,
    color = "grey35",
    linewidth = 0.75
  ) +
  geom_label(
    data = serotype_box_df,
    aes(x = x_mid, y = y, label = serotype_label),
    inherit.aes = FALSE,
    vjust = -1.65,
    fontface = "bold",
    size = 7,
    label.size = 0.35,
    label.padding = unit(0.16, "lines"),
    fill = "white"
  ) +
  geom_text(
    data = genotype_label_df,
    aes(x = x_mid, y = Inf, label = genotype_label),
    inherit.aes = FALSE,
    vjust = -0.35,
    fontface = "bold",
    size = 5
  ) +
  scale_fill_manual(values = serotype_colors, name = "Serotype") +
  shared_x +
  major_lineage_y +
  coord_cartesian(clip = "off") +
  labs(
    y = major_lineage_y_label
  ) +
  shared_plot_theme +
  theme(
    plot.margin = margin(34, 8, 12, 8)
  )

p_major_lineage_samples_aligned <- ggplot(
  lineage_bar_df,
  aes(x = x_index, y = n, fill = Continent)
) +
  geom_col(color = "black", width = 0.72, linewidth = 0.55) +
  geom_vline(
    data = genotype_label_df,
    aes(xintercept = xmin),
    inherit.aes = FALSE,
    color = "grey82",
    linewidth = 0.35
  ) +
  geom_vline(
    data = serotype_boundary_df,
    aes(xintercept = xmin),
    inherit.aes = FALSE,
    color = "grey35",
    linewidth = 0.75
  ) +
  scale_fill_manual(values = continent_colors, drop = FALSE) +
  scale_y_continuous(
  limits = c(0, sample_y_max),
  breaks = seq(0, sample_y_max, by = 50),
  labels = function(x) ifelse(x %% 100 == 0, x, ""),
  expand = expansion(mult = c(0, 0.08))
) +
  shared_x +
  labs(y = "N") +
  shared_plot_theme

p_major_lineage_temporal_aligned <- ggplot(
  lineage_presence_years,
  aes(x = x_index, y = year, color = Continent, group = interaction(x_index, Continent, run_id))
) +
  geom_line(linewidth = 0.55, alpha = 0.9) +
  geom_point(size = 0.75, alpha = 0.95) +
  geom_vline(
    data = genotype_label_df,
    aes(xintercept = xmin),
    inherit.aes = FALSE,
    color = "grey82",
    linewidth = 0.35
  ) +
  geom_vline(
    data = serotype_boundary_df,
    aes(xintercept = xmin),
    inherit.aes = FALSE,
    color = "grey35",
    linewidth = 0.75
  ) +
  scale_color_manual(values = continent_colors, drop = FALSE, guide = "none") +
  scale_y_continuous(
    limits = c(full_start_year, full_end_year),
    breaks = seq(full_start_year, full_end_year, by = 10),
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  shared_x +
coord_cartesian(clip = "off") +
labs(
  x = "Major lineage",
  y = "Yearly sampling",
  color = "Continent"
) +
  theme_classic(base_size = 20) +
  theme(
  plot.title = element_text(
    size = 20,
    face = "bold",
    hjust = 0.5
  ),

  axis.title.x = element_text(
    size = 18,
    face = "bold"
  ),

  axis.title.y = element_text(
    size = 18,
    face = "bold"
  ),

  axis.text.x = element_text(
    size = 18,
    angle = 90,
    hjust = 0.45,
    vjust = 0.3
  ),

  axis.text.y = element_text(
    size = 18
  ),

  axis.ticks.x = element_line(
    linewidth = 0.4
  ),

  axis.ticks.y = element_line(
    linewidth = 0.4
  ),

  legend.title = element_text(
    size = 16,
    face = "bold"
  ),

  legend.text = element_text(
    size = 16,
    face = "bold"
  ),

  legend.position = "bottom",

  plot.margin = margin(12, 8, 90, 8)
)

p_major_lineage_aligned <- p_major_lineage_rate_aligned /
  p_major_lineage_samples_aligned /
  p_major_lineage_temporal_aligned +
  plot_layout(heights = c(2.35, 0.8, 1.9), guides = "collect") &
  theme(legend.position = "bottom")

c
```

# Summary figure: serotype, genotype, and major lineage rates in one log-scale plot


```{r}
fig1 <- patchwork::wrap_elements(full = patchwork::patchworkGrob(p_serotype_genotype_with_samples)) /
  patchwork::wrap_elements(full = patchwork::patchworkGrob(p_major_lineage_aligned)) +
  patchwork::plot_layout(heights = c(1, 1.6))

ggsave(
  filename = "Figure1.svg",
  plot = fig1,
  width = 25 * (2/3), height = 25 * (2/3), units = "in",
  device = svglite::svglite
)
```
```{r}
ggsave(
  filename = "Figure2.svg",
  plot = p_serotype_by_protein,
  width = 25 * (2/3),
  height = 8 * (2/3),
  units = "in",
  device = svglite::svglite
)
```




